---
title: "DATA-643 Assignment - 02"
author:
- Mohamed Elmoudni
- Shazia Khan
- Senthil Dhanapal
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---


```{r, echo = FALSE, warning=FALSE, message=FALSE}


if (!require("ggplot2",character.only = TRUE)) (install.packages("ggplot2",repos = "http://cran.us.r-project.org", dep=TRUE))
if (!require("recommenderlab",character.only = TRUE)) (install.packages("ggplot2",repos = "http://cran.us.r-project.org", dep=TRUE))
if (!require("reshape2",character.only = TRUE)) (install.packages("ggplot2",repos = "http://cran.us.r-project.org", dep=TRUE))
if (!require("knitr",character.only = TRUE)) (install.packages("ggplot2",repos = "http://cran.us.r-project.org", dep=TRUE))

library(recommenderlab)
library(reshape2)
library(ggplot2)
library(knitr)

```


\newpage



#1 Requirements:

For project 2, you're asked to take some recommendation data (such as your
toy movie dataset, MovieLens, or another dataset of your choosing), and
implement at least two different recommendation algorithms on the data. For
example, content-based, user-user CF, and/or itemitem CF are the three
methods that we've covered in the course to date. You should evaluate
different approaches, using different algorithms, normalization techniques,
similarity methods, neighborhood sizes, etc. You don't need to be
exhaustive-these are just some suggested possibilities. You may use
whatever third party libraries you want. You should definitely tackle (at
least) accuracy metrics. Please provide at least one graph, and a textual
summary of your evaluation. 




#2 Data Load \newline



In this project we will be using the MovieLense dataset that comes with the recommenderlab package. 


```{r, echo = FALSE, warning=FALSE, message=FALSE}

set.seed(1)
data("MovieLense")


```



#3 Data Exploration \newline


In this phase we will perform data analysis and summary of the main characteristics of a dataset. We will be describing the data by means of statistical and visualization techniques


```{r, echo = FALSE, warning=FALSE, message=FALSE}
# Just look at first few lines of this file


tr <- (as(MovieLense,'data.frame'))
kable(head(tr))

############
kable(summary(tr), caption ="Data Summary")


kable(as.data.frame(table(tr$rating)), caption = "Ratings Frequency Table")

#qplot(tr$rating, tr$user)

qplot(tr$rating)
```


##3.1 Utility Matrix

```{r, echo = FALSE, warning=FALSE, message=FALSE}

r <- MovieLense
#slotNames(r)

vector_ratings <- as.vector(r@data)
unique(vector_ratings)

# we will assume that a rating of 0 represents a missing value, so we can remove them from vector_ratings:
vector_ratings <- vector_ratings[vector_ratings != 0]
table_ratings <- table(vector_ratings)
kable(data.frame(table_ratings))


# Visualizing a sample 
image(sample(r, 500), main = "Raw ratings")



```

##3.2 Data visualization

```{r, echo = FALSE, warning=FALSE, message=FALSE}
##########################
# Visualizing ratings
qplot(getRatings(r), binwidth = 1, 
      main = "Histogram of ratings", xlab = "Rating")

summary(getRatings(r))

###########################
# How about after normalization... "It looks better" 
qplot(getRatings(normalize(r, method = "Z-score")),
      main = "Histogram of normalized ratings", xlab = "Rating") 
summary(getRatings(normalize(r, method = "Z-score")))

###############################

# How many movies did people rate on average
qplot(rowCounts(r), binwidth = 10, 
      main = "Movies Rated on average", 
      xlab = "# of users", 
      ylab = "# of movies rated")


###############


# What is the mean rating of each movie
qplot(colMeans(r), binwidth = .1, 
      main = "Mean rating of Movies", 
      xlab = "Rating", 
      ylab = "# of movies")


```



#4 Model Evaluation


In this section, we will evaluate IBCF and UBCF using Cosine, Jaccard, and Pearson methods. 

```{r, echo=FALSE , message=FALSE, warning=FALSE}
#df <- data.frame(name=character(),RMSE=numeric,MSE=numeric,MAE=numeric)


EvaluationSummary <- function(acc_matrix, methodName)
{
  df1<-as.data.frame(apply(acc_matrix,2,mean,na.rm=T))
  names(df1) <- methodName 
  return(df1)
}

```

Split dataset into Training and Testing

```{r, echo=FALSE , message=FALSE, warning=FALSE}

ratings_movies <- MovieLense
percentage_training <- 0.8
items_to_keep <- 10
rating_threshold <- 3
eval_sets <- evaluationScheme(data = ratings_movies, method = "split",
                              train = percentage_training, 
                              given = items_to_keep, 
                              goodRating =rating_threshold, k = NULL) 



```


##4.1 UBCF



UBCF: In user-based collaborative filtering predictions are based on aggregated ratings from the closest users (nearest neighbors).  Nearest neighbors are defined based on similarity between users, which is calculated using available ratings. This method works under the assumption that the users with similar ratings will rate items similarly.
In this case we will try UBCF using Cosine,Jaccard, and Pearson.


###4.1.1 UBCF Cosine 


```{r, echo=FALSE , message=FALSE, warning=FALSE}

rec <- Recommender(data = getData(eval_sets, "train"),
                                method = 'UBCF', 
                                param=list(normalize = "Z-score",method="Cosine"))

pred <- predict(object = rec, 
                           newdata =getData(eval_sets, "known"), 
                           n = 10, type = "ratings")

acc <- calcPredictionAccuracy(x = pred, 
                                        data = getData(eval_sets, "unknown"), 
                                        byUser =TRUE)

head(acc)

dfEvalSummary <- EvaluationSummary(acc,'UBCF-Cosine')





```


###4.1.2 UBCF Jaccard


```{r, echo=FALSE , message=FALSE, warning=FALSE}

rec <- Recommender(data = getData(eval_sets, "train"),
                                method = 'UBCF', 
                                param=list(normalize = "Z-score",method="Jaccard"))

pred <- predict(object = rec, 
                           newdata =getData(eval_sets, "known"), 
                           n = 10, type = "ratings")

acc <- calcPredictionAccuracy(x = pred, 
                                        data = getData(eval_sets, "unknown"), 
                                        byUser =TRUE)

head(acc)

dfEvalSummary <- cbind(dfEvalSummary,EvaluationSummary(acc,'UBCF-Jaccard'))





```

###4.1.3 UBCF Pearson



```{r, echo=FALSE , message=FALSE, warning=FALSE}

rec <- Recommender(data = getData(eval_sets, "train"),
                                method = 'UBCF', 
                                param=list(normalize = "Z-score",method="Pearson"))

pred <- predict(object = rec, 
                           newdata =getData(eval_sets, "known"), 
                           n = 10, type = "ratings")

acc <- calcPredictionAccuracy(x = pred, 
                                        data = getData(eval_sets, "unknown"), 
                                        byUser =TRUE)

head(acc)

dfEvalSummary <- cbind(dfEvalSummary,EvaluationSummary(acc,'UBCF-Pearson'))





```


## 4.2 IBCF


Item-based collaborative filtering where we consider set of movies rated by the user and computes item similarities with the targeted movie. Once similar movies are found, and then rating for the new movie is predicted by taking weighted average of the user's rating on these similar movies.
in this case we will try IBCF using Cosine,Jaccard, and Pearson.



###4.2.1 IBCF Cosine 




```{r, echo=FALSE , message=FALSE, warning=FALSE}

# IBCF Cosine

rec <- Recommender(data = getData(eval_sets, "train"),
                                method = 'IBCF', 
                                param=list(normalize = "Z-score",method="Cosine"))

pred <- predict(object = rec, 
                           newdata =getData(eval_sets, "known"), 
                           n = 10, type = "ratings")

acc <- calcPredictionAccuracy(x = pred, 
                                        data = getData(eval_sets, "unknown"), 
                                        byUser =TRUE)

head(acc)

dfEvalSummary <- cbind(dfEvalSummary,EvaluationSummary(acc,'IBCF-Cosine'))

```


###4.2.2 IBCF Jaccard


```{r, echo=FALSE , message=FALSE, warning=FALSE}

rec <- Recommender(data = getData(eval_sets, "train"),
                                method = 'IBCF', 
                                param=list(normalize = "Z-score",method="Jaccard"))

pred <- predict(object = rec, 
                           newdata =getData(eval_sets, "known"), 
                           n = 10, type = "ratings")

acc <- calcPredictionAccuracy(x = pred, 
                                        data = getData(eval_sets, "unknown"), 
                                        byUser =TRUE)

head(acc)

dfEvalSummary <- cbind(dfEvalSummary,EvaluationSummary(acc,'IBCF-Jaccard'))

```




###4.2.3 IBCF Pearson



```{r, echo=FALSE , message=FALSE, warning=FALSE}

rec <- Recommender(data = getData(eval_sets, "train"),
                                method = 'IBCF', 
                                param=list(normalize = "Z-score",method="Pearson"))

pred <- predict(object = rec, 
                           newdata =getData(eval_sets, "known"), 
                           n = 10, type = "ratings")

acc <- calcPredictionAccuracy(x = pred, 
                                        data = getData(eval_sets, "unknown"), 
                                        byUser =TRUE)

head(acc)

dfEvalSummary <- cbind(dfEvalSummary,EvaluationSummary(acc,'IBCF-Pearson'))


```


#5 Calculate Average of all Errors


Below we will be calculating averages for both IBCF and UBCF for all three methods (Cosine, Jaccard, and Pearson)



```{r, echo=FALSE , message=FALSE, warning=FALSE}

avg <- apply(as.matrix(dfEvalSummary),2,mean)

df1 <- as.data.frame(t(avg))
rownames(df1) <- 'AverageOfAllErrors'
kable(rbind(dfEvalSummary,df1), caption = 'Evaluation Summary')

```



#6 Conclusion


From the Evaluation Summary table above, UBCF Cosine shows the least average error for RMSE, MSE, and MAE compared with other methods. Therefore, UBCF-Cosine is the choosen method for recommendation. 
Next we will run the recommendation based on the best chosen method - UBCF Cosine- and show the result of one test user before and after prediction:


```{r, echo=FALSE , message=FALSE, warning=FALSE}

final_rec <- Recommender(data = MovieLense,
                                method = 'UBCF', 
                                param=list(normalize = "Z-score",method="Cosine"))

pred <- predict(object = final_rec, newdata =MovieLense, n = 10, type = "ratings")

m1 <- head(as(pred,'matrix'))
m1 <- m1['2',2:4,drop=FALSE]

m <- as(MovieLense,'matrix')
m2 <- m['2',2:4,drop=FALSE]

df2 <- rbind(as.data.frame(m2), as.data.frame(m1))
rownames(df2) <- c('Before','After')
kable(df2, caption ='Test Using UBCF Cosine')

```


\newpage

#Appendix A: DATA643 Assignment 02 R Code 
```{r code=readLines(knitr::purl('https://raw.githubusercontent.com/simonnyc/IS-643/master/Assignment02/Assignment02_08.Rmd', documentation = 0)), eval = FALSE}

```


