---
title: "Assignment01"
author: "Mohamed Elmoudni"
date: "February 12, 2017"
output: pdf_document
---

## Project 1 : 

Requirements:

The goal of this assignment is to help you build your intuition about recommender systems, with a basic soup to nuts implementation coded "from scratch." 
 
Your task is to build a very basic recommender system, first by writing your own functions, then by replacing those functions with those provided in an R Package or a Python library (such as scikitlearn). 
 
 - You should very briefly first describe the recommender system that you're going to build out from a business perspective, e.g. "This system recommends movies to users." 
 
 - You can find a dataset, or build out your own toy dataset and load into (for example) an R or pandas dataframe, a Python dictionary or list of lists, (or other data structure of your choosing). 

 - You can use either collaborative filtering, or a hybrid of content management and collaborative filtering.   

 - You are encouraged to hand code at least your similarity function. 
 - After you have built out your own code base, create an alternate version using packages or libraries.  Compare the results and performance. 

 - You are also encouraged to think about how to best handle missing data. 
 
 - Your code should be turned in an RMarkdown file or a Jupyter notebook, and posted to Github. You may work in a small group (2 or 3 people) on this assignment.  While you're never discouraged from adding features or advanced capabilities such as regularization and matrix factorization methods, it is not expected at this point in the course.   


```{r}
library(recommenderlab)
library(reshape2)
library(ggplot2)
library(knitr)
######################################

# Read training file along with header
dt<-read.csv("https://raw.githubusercontent.com/simonnyc/IS-643/master/Assignment01_data.csv", header=TRUE)

#############
# Data exploration


# Just look at first few lines of this file
kable(head(dt))

```

Data Exploration

```{r}

# Summary
kable(summary(dt), caption ="Data Summary")

#frequency table

table(dt$Ratings)
qplot(dt$Ratings, dt$User)

qplot(dt$Ratings)

#[head(sort(dt$Ratings,decreasing=TRUE), n = 5)]

#head(dt[sort(dt$Ratings, decreasing=TRUE), ], 100)

```


Implementing Item based recommender systems, like user based collaborative filtering, requires two steps:

1. Calculating Item similarities
2. Predicting the targeted item rating for the targeted User.

Step1: Calculating Item Similarity:
we calculate the similarity between co-rated items. We use cosine similarity or pearson-similarity to compute the similarity between items. 
The output for step is similarity matrix between Items.


```{r, echo=FALSE}
#step 1: item-similarity calculation co-rated items are considered and similarity between two items
#are calculated using cosine similarity

g<-acast(dt, User ~ Items, value.var = 'Ratings')
#g[is.na(g)] = 0
df<- data.frame(g)
df$userno<- as.numeric(rownames(df))

#re-arrange the columns
library(dplyr)
df <- df %>%
  select(item_10:item_12, everything())

df <- df %>%
  select(item_1:item_9, everything())

df <- df %>%
  select(userno, everything())




#install.packages("lsa")
library(lsa)
x = df[,2:ncol(df)]
x[is.na(x)] = 0
itemSimil = cosine(as.matrix(x))

kable(itemSimil)

```


Step2: Predicting the targeted item rating for the targeted User 

Recommending Top N items:
Once all the non rated movies are predicted we recommend top N movies to a user Code for Item based collaborative filtering in R:

```{r}

ratings<- as.matrix(x)
userRecmd <- function(userno)
{

userRatings <- ratings[userno,]
#

non_rated_items <- names(userRatings[is.na(userRatings)])
rated_items <- names(userRatings[!is.na(userRatings)])
m1 <- itemSimil[non_rated_items,]
v1 <- apply(m1,1,function(x) sum((x*ratings[userno,]),na.rm = T)/(sum(x[rated_items])))
ratings[userno,names(v1)] <- v1
#

return(ratings[userno,])
}

```

####  Recommend for User 10
Now we will use the above to find a rating for user 10

```{r}
userRecmd(10)
```

str(x)
```{r}
################################## Mohamed Changes 
x = df[,2:ncol(df)]
#x[is.na(x)] = 0
x<- as.matrix(x)
x <- as(x, "realRatingMatrix")
kable(as(x, "matrix"))

rec=Recommender(x[1:nrow(x)],method="UBCF", param=list(normalize = "Z-score",method="Cosine", nn=12))
rec2 <- predict(rec, x[1:nrow(x)])
(as(rec2, "matrix")['10',])

############################
rec=Recommender(x[1:nrow(x)],method="UBCF", param=list(normalize = "Z-score",method="Jaccard", nn=5))
rec2 <- predict(rec, x[1:nrow(x)])
as(rec2, "matrix")['10',]
############################
rec=Recommender(x[1:nrow(x)],method="UBCF", param=list(normalize = "Z-score",method="Jaccard"))
rec2 <- predict(rec, x[1:nrow(x)])
as(rec2, "matrix")['10',]

##################

rec=Recommender(x[1:nrow(x)],method="IBCF", param=list(normalize = "Z-score",method="Cosine"))
rec2 <- predict(rec, x[1:nrow(x)])
as(rec2, "matrix")['10',]

##################

```





```{r}

```

```{r}

```








