---
title: "Assignment01"
author: "Mohamed Elmoudni"
date: "February 12, 2017"
output: pdf_document
---

## Project 1 : 

Requirements:

The goal of this assignment is to help you build your intuition about recommender systems, with a basic soup to nuts implementation coded "from scratch." 
 
Your task is to build a very basic recommender system, first by writing your own functions, then by replacing those functions with those provided in an R Package or a Python library (such as scikitlearn). 
 
 - You should very briefly first describe the recommender system that you're going to build out from a business perspective, e.g. "This system recommends movies to users." 
 
 - You can find a dataset, or build out your own toy dataset and load into (for example) an R or pandas dataframe, a Python dictionary or list of lists, (or other data structure of your choosing). 

 - You can use either collaborative filtering, or a hybrid of content management and collaborative filtering.   

 - You are encouraged to hand code at least your similarity function. 
 - After you have built out your own code base, create an alternate version using packages or libraries.  Compare the results and performance. 

 - You are also encouraged to think about how to best handle missing data. 
 
 - Your code should be turned in an RMarkdown file or a Jupyter notebook, and posted to Github. You may work in a small group (2 or 3 people) on this assignment.  While you're never discouraged from adding features or advanced capabilities such as regularization and matrix factorization methods, it is not expected at this point in the course.   

#####################################

Our recommender system recommends movies based on content based recommendation system. It uses a m*n matrix with users as rows and movies as columns. Our function predicts the missing ratings for a movie based on the similar movies. Similarity of movies is calculated based on cosine of vectors method. It uses that cosine values as weights to the ratings of other movies to predict the missing ratings.

We used a toy dataset to train our functions and used movie dataset downloaded from online to test the functions.

We have also tested using Recommenderlab, which is one of the R packages. Even though our values were not matching exactly, we got values somewhat closer.



```{r}
library(recommenderlab)
library(reshape2)
library(ggplot2)
library(knitr)
######################################

# Read training file along with header
dt<-read.csv("https://raw.githubusercontent.com/simonnyc/IS-643/master/Assignment01_data.csv", header=TRUE)

#############
# Data exploration


# Just look at first few lines of this file
kable(head(dt))

```

Data Exploration

```{r}

# Summary
kable(summary(dt), caption ="Data Summary")

#frequency table

table(dt$Ratings)
qplot(dt$Ratings, dt$User)

qplot(dt$Ratings)

#[head(sort(dt$Ratings,decreasing=TRUE), n = 5)]

#head(dt[sort(dt$Ratings, decreasing=TRUE), ], 100)

```


Implementing Item based recommender systems, like user based collaborative filtering, requires two steps:

1. Calculating Item similarities
2. Predicting the targeted item rating for the targeted User.

Step1: Calculating Item Similarity:
we calculate the similarity between co-rated items. We use cosine similarity compute the similarity between items. 

The output for step is similarity matrix between Items.


```{r, echo=FALSE}
#step 1: item-similarity calculation co-rated items are considered and similarity between two items
#are calculated using cosine similarity

g<-acast(dt, User ~ Items, value.var = 'Ratings')
#g[is.na(g)] = 0
df<- data.frame(g)
df$userno<- as.numeric(rownames(df))

#re-arrange the columns
library(dplyr)
df <- df %>%
  select(item_10:item_12, everything())

df <- df %>%
  select(item_1:item_9, everything())

df <- df %>%
  select(userno, everything())




#install.packages("lsa")
library(lsa)
x = df[,2:ncol(df)]
x[is.na(x)] = 0
itemSimil = cosine(as.matrix(x))

kable(itemSimil)

```


Step2: Predicting the targeted item rating for the targeted User using content based system


```{r}
ratings<- as.matrix(x)
userRecmd <- function(userno)
{

  userRatings <- ratings[userno,]
  userRatings[userRatings==0] <- NA
  #
  
  non_rated_items <- names(userRatings[is.na(userRatings)])
  rated_items <- names(userRatings[!is.na(userRatings)])
  m1 <- itemSimil[non_rated_items,]
  v1 <- apply(m1,1,function(x) sum((x*ratings[userno,]),na.rm = T)/(sum(x[rated_items])))
  #ratings[userno,names(v1)] <- v1
  #

  return(v1)
}

```

####  Recommend for User 10

Following line predicts the missing ratings for User 10

```{r}
userRecmd(10)
```


User defined function for user based collobarative filtering

```{r}

Recmd.UIB <- function(M, user, k)
{
  #M <- ratings
  #user <- 'u10'
  M[is.na(M)] <- 0
  rowsums <- rowSums(M)
  rowcounts <- apply(M,1,function(x) length(x[x>0]))
  rowavg <- rowsums/rowcounts
  for(r in 1:nrow(M))
  {
    for(c in 1:ncol(M))
    {
      if(M[r,c] >0 )
      {
        M[r,c] <- M[r,c] - rowavg[r]
      }
    }
  }
  
  user.M <- M[user,]
  if(is.element(0,user.M)==F) 
  {
    print('no missing ratings')
    return()
  }
  
  otherusers.M <- M[rownames(M)!=user,]
  sim.M <- matrix(,nrow(otherusers.M))
  rownames(sim.M) <- rownames(otherusers.M)
  
  for(r in 1:nrow(otherusers.M))
  {
    sim.user <- rownames(otherusers.M)[r]
    sim.M[sim.user,1] <- cosine(user.M,otherusers.M[r,])
  }
  sim.M[is.na(sim.M)] <- -1
  sim.M <- sim.M[order(-sim.M[,1]),,drop=F]
  
  non.rated.items <- names(user.M[user.M==0])
  
  #top k similar users who have rated the items not rated by the current user
  if(k > nrow(sim.M)) 
  {
    k <- nrow(sim.M)
  }
  sim.user.k <- sim.M[1:k,,drop=F]
  sim.user.k1 <- ratings[rownames(sim.user.k),non.rated.items,drop=F]
  sim.user.k1 <- cbind(sim.user.k1,sim=sim.user.k)
  colnames(sim.user.k1) <- c(non.rated.items,'sim')
  sim.user.k2 <- matrix(,ncol=ncol(sim.user.k1))
  for(r in 1:nrow(sim.user.k1))
  {
    v.temp <- sim.user.k1[r,,drop=F]
    if(is.element(NA,v.temp)!=T)
    {
      sim.user.k2 <- rbind(sim.user.k2,v.temp)
      
    }
  }
  sim.user.k2 <- sim.user.k2[2:nrow(sim.user.k2),]
  sim.user.k3 <- sim.user.k2[,1:ncol(sim.user.k2)-1]
  sim.user.k4 <- sim.user.k2[,ncol(sim.user.k2),drop=F]
  
  v1 <- apply(sim.user.k3,2,function(x) sum(x*sim.user.k4)/sum(sim.user.k4))
  return(v1)
  
  
}


```

Prediction for same user 10 using User based collobarative filtering

```{r}

Recmd.UIB(ratings,'10',5)

```


Following lines predict the ratings for the user 10 using built-in package :- RecommenderLab package and using Item based collobarative filtering
```{r}

m <- ratings
m[m==0] <- NA

affinity.matrix<- as(m,"realRatingMatrix")
Rec.model<-Recommender(affinity.matrix,method="IBCF", 
                       param=list(normalize = "Z-score",method="Cosine"))

u <- predict(Rec.model, affinity.matrix[10,], type="ratings")
as(u, "list")

```




Following lines predict the ratings for the user 10 using built-in package :- RecommenderLab package and using User based collobarative filtering
```{r}

m <- ratings
m[m==0] <- NA

affinity.matrix<- as(m,"realRatingMatrix")
Rec.model<-Recommender(affinity.matrix,method="UBCF", 
                       param=list(normalize = "Z-score",method="Cosine"))

u <- predict(Rec.model, affinity.matrix[10,], type="ratings")
as(u, "list")

```




```{r}


################################## Mohamed Changes 
x = df[,2:ncol(df)]
#x[is.na(x)] = 0
x<- as.matrix(x)
x <- as(x, "realRatingMatrix")
kable(as(x, "matrix"))

rec=Recommender(x[1:nrow(x)],method="UBCF", param=list(normalize = "Z-score",method="Cosine", nn=12))
rec2 <- predict(rec, x[1:nrow(x)])
(as(rec2, "matrix")['10',])

############################
rec=Recommender(x[1:nrow(x)],method="UBCF", param=list(normalize = "Z-score",method="Jaccard", nn=5))
rec2 <- predict(rec, x[1:nrow(x)])
as(rec2, "matrix")['10',]
############################
rec=Recommender(x[1:nrow(x)],method="UBCF", param=list(normalize = "Z-score",method="Jaccard"))
rec2 <- predict(rec, x[1:nrow(x)])
as(rec2, "matrix")['10',]

##################

rec=Recommender(x[1:nrow(x)],method="IBCF", param=list(normalize = "Z-score",method="Cosine"))
rec2 <- predict(rec, x[1:nrow(x)])
as(rec2, "matrix")['10',]

##################

```




